\vspace{-.5cm} Lights: a generalized joint model for high-dimensional multivariate longitudinal data and censored durations \vspace{.5cm}

### 2022-02-28 manuscript/paper.tex
update 1

\newcommand{\dd}{\mathrm{d}}
\newcommand{\ind}[1]{\mathds{1}_{#1}}
\newcommand{\norm}[1]{\|#1\|}
\newcommand{\cM}{\mathcal M}
\newcommand{\cD}{\mathcal D}
\newcommand{\cC}{\mathcal C}
\newcommand{\cG}{\mathcal G}
\newcommand{\cN}{\mathcal N}
\newcommand{\cA}{\mathcal A}
\newcommand{\sA}{\mathscr A}
\newcommand{\cQ}{\mathcal Q}
\newcommand{\cU}{\mathcal U}
\newcommand{\cR}{\mathcal R}
\newcommand{\cH}{\mathcal H}
\newcommand{\cB}{\mathcal B}
\newcommand{\cS}{\mathcal S}
\newcommand{\cI}{\mathcal I}
\newcommand{\R}{\mathds R}
\newcommand{\N}{\mathds N}
\newcommand{\E}{\mathds E}
\renewcommand{\P}{\mathds P}
\newcommand{\bSigma}{\textbf{$\Sigma$}}

@@

This paper introduces a prognostic method called \textit{lights} to deal with the problem of joint modeling of longitudinal data and censored durations, where a large number of both longitudinal and time-independent features are available. Yet there is no standard model so far to learn from such high-dimensional multivariate longitudinal data in a survival analysis setting.
Shared association features are learned from the longitudinal processes and included as potential risk factor in a group-specific Cox model. Appropriate penalties are then used during inference to allow flexibility in modeling the dependency between the longitudinal features and the event time.
The statistical performance of the method is examined on an extensive Monte Carlo simulation study, and finally illustrated on a publicly available dataset.
Our proposed method significantly outperforms the state-of-the-art survival models regarding risk prediction in terms of C-index, with a computing time orders of magnitude faster. In addition, it provides powerful interpretability by automatically pinpointing significant features being relevant from a practical perspective. Thus, we propose a powerfull tool with the ability of automatically determining significant prognostic longitudinal features, which is of increasing importance in many areas: for instance personalized medicine, or churn prediction in a customer profile and activity monitoring setting, to name but a few.\\

@@

In this paper, we propose a method called \textit{lights} (generaLized joInt hiGH-dimensional longiTudinal Survival) which is from both JLCMs and SREMs, since we also include features extracted from the longitudinal processes as potential risk factor in the survival model, which is a group-specific Cox model~\citep{Cox1972JRSS} with high-dimensional shared associations.
To allow flexibility in modeling the dependency between the longitudinal features and the event time, we use appropriate penalties : elastic net~\citep{zou2005regularization}
for time-independent feature selection in the latent class membership, and sparse group lasso~\citep{simon2013sparse} in the survival model : this penalty acts like the lasso at the trajectory level, namely an entire trajectory may drop out of the model, and yields sparsity for a given trajectory.
Inference is achieved using an efficient and novel proximal Quasi-Newton Expectation Maximization algorithm (prox-QNEM).
Posterior estimates of the latent class membership probabilities given the longitudinal process and the fact that the subject is still alive at prediction time -- in a so called ``real-time prediction paradigm'' -- are then used as discriminative marker rule for risk prediction in the cross-validation procedure for selecting the best regularization hyper-parameters.
Hence, the method provides interpretations of the high-dimensional longitudinal features, thus offering a powerful tool for real-time decision support in a customer's satisfaction monitoring context for instance, or else for clinical decision making in patient monitoring.

@@

@@

To quantify the effect of the longitudinal outcomes on the risk for an event, we use a Cox~\citep{Cox1972JRSS} relative risk model of the form
\begin{equation}
  \label{eq:intensity-model}
  \lambda(t|\cM_k(t), G = k) = \lambda_0(t) \exp \Big\{\sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \Psi_a^l \Big\},
\end{equation}
where $\lambda_0(\cdot)$ is an unspecified baseline hazard function.

For each $l$-th longitudinal outcome, we consider $\cA \in \N_+$ known features $\Psi_a$ defining a shared association with $\gamma_{k,a}^l \in \R^{\cA}$ the corresponding joint association parameters.  These features are extracted from $Y$ by the \texttt{Python} library \texttt{tsfresh}. This can be viewed as a generalization of SREMs~\citep{rizopoulos2010jm}. Let us finally denote
\[\gamma_k= (\gamma_{k,1}^1, \ldots, \gamma_{k,\cA}^1, \ldots, \gamma_{k,1}^L, \ldots, \gamma_{k,\cA}^L )^\top \in \R^{L\cA}, \] so that
\begin{equation}
  \label{eq:def-asso-features}
  \lambda(t|\Psi, G = k) = \lambda_0(t) \exp \big\{\gamma_k^\top \Psi \big\}.
\end{equation}

We suppose a group-specific marker trajectory and a generalized linear mixed model for each longitudinal marker given subgroup $G$, so that for the $l$-th outcome at time $t \geq 0$ one has
\begin{equation}
	\label{eq:link-function}
 	h_l\big(\E[Y^l(t)|b^l, G=k]\big) = m_k^l(t)
 \end{equation}
where $h_l$ denotes a known one-to-one link function, and $m_k^l$ the linear predictor such that
\[ m_k^l(t) = u^l(t)^\top\beta_k^l + v^l(t)^\top b^l \]
where $u^l(t) \in \R^{q_l}$ is a row vector of (possibly) time-varying features with corresponding unknown fixed effect parameters $\beta_k^l$, and $v^l(t) \in \R^{r_l}$ is a row vector of (possibly) time-varying features with $r_l \leq q_l$ and where the corresponding subject-and-longitudinal outcome specific random effects $b^l$ that does not depend on the group membership, which is not a strong modeling assumption.
\begin{assumption}
\label{assumption1}
We suppose that the random effects are conditional independent of the group membership on the association features, and that the latter remain independent conditional on the observed data (namely $T$, $\Delta$ and $Y$).
\end{assumption}
A suitable distributional assumption for the random effects component is a zero-mean multivariate normal distribution~\citep{hickey2016joint}, that is
\begin{equation}
   \label{eq:random-effect-law}
   b^l \sim \cN(0, D_{ll})
 \end{equation} 
with $D_{ll} \in \R^{r_l \times r_l}$ the unstructured variance-covariance matrix. 
To account for dependence between the different longitudinal outcome types, we let $\text{Cov}[b^l,b^{l'}] = D_{ll'}$ for $l \ne l'$ and we denote
\[ D = 
\begin{bmatrix}
  D_{11} & \cdots & D_{1L}\\
  \vdots &  \ddots & \vdots \\
  D_{1L}^\top & \cdots & D_{LL}
\end{bmatrix}
\]
the global variance-covariance matrix.
In the sequel of the paper, our aim is to associate the true and unobserved value $m_k^l(t)$ of the $l$-th longitudinal outcome at time $t$ with the event outcome $T^\star$.

\subsection{Graphical model for the lights}
Figure~\ref{fig:graphical-model} below gives the graphical representation for the lights model to get a beter view of the dependence structure for the model.
\begin{figure}[!htb]
\centering
\resizebox{.9\textwidth}{!}{
\begin{tikzpicture}
\hspace{3cm}
\tikzstyle{main}=[circle, minimum size = 18mm, ultra thick, draw =black!80, node distance = 13mm,text centered]
\tikzstyle{var}=[rectangle, minimum size = 16mm, ultra thick, draw =black!80, node distance = 13mm,text centered]
\tikzstyle{connect}=[-latex, thick]
\tikzstyle{box}=[rectangle, draw=black!100]
  \node[var, fill = white!100] (beta) [] {\Large$\bm{\beta_k^l}$};
  \node[var, fill = white!100] (xi) [below=4cm of beta] {\Large$\bm{\xi_k}$};
  \node[var, fill = white!100] (gamma) [below=3cm of xi] {\Large$\bm{\gamma_k}$};
  \node[main, fill = white!100] (G) [right=4cm of xi] {\Large$\bm{G_i}$};
  \node[main, fill = white!100] (mu) [right=4cm of beta] {\Large$\bm{M_{ik}^l(t)}$};
  \node[main, fill = black!30] (Y) [right=of mu] {\Large$\bm{Y_i^l(t)}$};
  \node[main, fill = white!100] (b) [above=of mu] {\Large$\bm{b_i^l}$};
  \node[var, fill = white!100] (phi) [above=of Y] {\Large$\bm{\phi_l}$};
  \node[var, fill = white!100] (Sigma) [left=of b] {\Large$\bm{D}$};
  \node[main, fill = black!30] (W) [below=3cm of Y] {\Large$\bm{\Psi_i}$};
  \node[main, fill = white!100] (lambda) [right=4cm of gamma] {\Large$\bm{\lambda_i(t)}$};
  \node[main, fill = black!30] (T) [right=of lambda] {\Large$\bm{T_i}$};
  \node[main, fill = black!30] (X) at (3, -8.3) {\Large$\bm{X_i}$};

@@

\end{tikzpicture}}
\caption{Graphical representation of the lights model. Square boxes denote observed data while circles denote unobserved (including random) terms. The red-dashed box indicates that the association features and the random effects are correlated, which, in addition to the link with the latent group membership, gives rise to the joint model.}
\label{fig:graphical-model}
\end{figure}

@@

Consider an independent and identically distributed (i.i.d.) cohort of $n$ subjects
\[ \cD_n = \big\{ (x_1, y_1^1, \ldots, y_1^L, t_1, \delta_1, \Psi_1), \ldots, (x_n, y_n^1, \ldots, y_n^L, t_n, \delta_n, \Psi_n) \big\} \]
where for each subject $i=1, \ldots, n$, process $Y_i^l$ is measured $n_i^l$ times at $t_{i1}^l, \ldots, t_{in_i^l}^l$ (which can differ between subjects and outcomes) with $t_{ij}^l \leq t_{ij+1}^l$ for all $j=1, \ldots, n_i^l-1$ and such that
\[ y_i^l=(y_{i1}^l, \ldots, y_{in_i^l}^l)^\top \in \R^{n_i^l} \quad \text{ with } \quad y_{ij}^l=Y_i^l(t_{ij}^l) \]
for all $l=1, \ldots, L$. 
For the $i$-th subject, let us denote 
\[
\left\{
    \begin{array}{ll}
        y_i &= ({y_i^1}^\top \cdots {y_i^L}^\top)^\top \in \R^{n_i},\\
        b_i &= ({b_i^1}^\top \cdots {b_i^L}^\top)^\top \in \R^r,
    \end{array}
\right.
\]
with $n_i = \sum_{l=1}^L n_i^l$ and $r = \sum_{l=1}^L r_l$ the total number of longitudinal measurements (for subject $i$) and the total dimension of the random effects respectively, as well as the following design matrices
\[ U_i = 
\begin{bmatrix}
  U_{i1} & \cdots & 0\\
  \vdots &  \ddots & \vdots \\
  0 & \cdots & U_{iL}
\end{bmatrix} 
\in \R^{n_i \times q}
\qquad \text{and} \qquad
V_i = 
\begin{bmatrix}
  V_{i1} & \cdots & 0\\
  \vdots &  \ddots & \vdots \\
  0 & \cdots & V_{iL}
\end{bmatrix}
\in \R^{n_i \times r}
\]
with $q = \sum_{l=1}^L q_l$ and where for all $l=1, \ldots, L$, one writes
\[
\left\{
    \begin{array}{ll}
        U_{il} &= \big(u_i^l(t_{i1}^l)^\top \cdots u_i^l(t_{in_i^l}^l)^\top\big)^\top \in \R^{n_i^l \times q_l},\\
        V_{il} &= \big(v_i^l(t_{i1}^l)^\top \cdots v_i^l(t_{in_i^l}^l)^\top\big)^\top \in \R^{n_i^l \times r_l}.
    \end{array}
\right.
\]
From now on, all computations are done conditionally on the design matrices $(U_i)_{i=1, \dots, n}$ and $(V_i)_{i=1, \dots, n}$.
\begin{assumption}
\label{indep-hyp-2}
Suppose that for a given subject $i$ and conditional on the random effects and the group membership, all $Y_i^l$ are independent for $l=1, \dots, L$.
\end{assumption}
Assumption~\ref{indep-hyp-2} is similar to standard modeling hypothesis, see for instance~\citet{tsiatis2004joint}.
For all $k = 0, \ldots, K-1$, we denote 
\[ \beta_k = ({\beta_k^1}^\top \cdots {\beta_k^L}^\top)^\top \in \R^q \] 
and 
\[M_{ik} = U_i\beta_k + V_ib_i \in \R^{n_i}. \]

@@

Let us denote 
\[ \theta = \big(\xi_0^\top \cdots \xi_{K-1}^\top, \beta_0^\top \cdots \beta_{K-1}^\top, \phi^\top, \text{vech}(D), \lambda_0^\top, \gamma_0^\top \cdots \gamma_{K-1}^\top\big)^\top \in \R^\vartheta \]
the collection of the $\vartheta \in \N_+$ unknown parameters to estimate. Note that we include $\lambda_0$ in $\theta$. Indeed, we will estimate the baseline using nonparametric maximum likelihood, as a function taking mass at each failure time. So in $\theta$, $\lambda_0$ represents the vector of these baseline hazard quantities, with a dimension equal to the number of unique failure times (denoted $J$ thereafter).
To write the log-likelihood $\ell_n(\theta)$ (rescaled by $n^{-1}$)  for samples in $\cD_n$, corresponding to the joint distribution of the time-to-event and longitudinal outcomes, let us make the following hypothesis.
\begin{assumption}
\label{indep-hyp-3}
Assume that the random effects vector $b_i$, the group membership account for the association between the longitudinal and event outcomes and the association features $\Psi_i$, that is
\begin{equation}
  \label{eq:ind-hyp}
  f(t_i, \delta_i, y_i | b_i, G_i = k, \Psi_i ; \theta) = f(t_i, \delta_i| G_i = k, \Psi_i ; \theta) f(y_i | b_i, G_i = k ; \theta)
\end{equation}
for all $i=1, \ldots, n$.
\end{assumption}
Assumption~\ref{indep-hyp-3} is a generalization of classical hypothesis used in SREMs and JLCMs~\citep{hickey2016joint}.
Then, one has
\begin{align}
	\label{eq:log-lik}
	\ell_n(\theta) &= \ell_n(\theta ; \cD_n) \nonumber \\
	&= n^{-1} \sum_{i=1}^n \log \sum_{k=0}^{K-1} \pi_{\xi_k}(x_i) \int_{\R^r} f(t_i, \delta_i| b_i, G_i = k, \Psi_i ; \theta) f(y_i | b_i, G_i = k, \Psi_i; \theta) f(b_i ; \theta) \dd b_i \\
  &= n^{-1} \sum_{i=1}^n \log \sum_{k=0}^{K-1} \pi_{\xi_k}(x_i) f(t_i, \delta_i| G_i = k, \Psi_i ; \theta) f(y_i | G_i = k; \theta) \dd b_i,
\end{align}
where 
\[f(t_i, \delta_i| G_i = k, \Psi_i ; \theta) = \big[\lambda(t_i|G_i = k, \Psi_i)\big]^{\delta_i} \exp \Big\{-\int_0^{t_i} \lambda(s|G_i = k, \Psi_i) \dd s \Big\}. \]

@@

In many practical applications, subjects show highly nonlinear longitudinal trajectories. In~\eqref{eq:log-lik}, the complete longitudinal history is required for the computation of the survival function. Hence, in order to produce a good estimate of $\cM_k(t)$, we consider a flexible representations for $u^l(t)$ using a high-dimensional vector of time monomials, namely 
\begin{equation}
  \label{eq:alpha-def}
  u^l(t) = (1, t, t^2, \ldots, t^\alpha)^\top
\end{equation}
with $\alpha \in \N_+$.
The idea here is to let the practitioner choose a suitable polynomial order $\alpha$ for the representation -- that could also be tuned automatically using a model selection procedure and AIC~\citep{akaike1974new} for instance.
We then let \[v^l(t) = (1, t)^\top\] so that each trajectory of each subject gets an affine random effect. Hence with this choice in practice, one has $q_l=\alpha + 1$ and $r_l=2$ for all $l=1, \ldots, L$.

@@

\subsection{A Quasi-Newton EM}
\label{sec:prox-QNEM}

>>>>

\newcommand{\dd}{\mathrm{d}}
\newcommand{\ind}[1]{\mathds{1}_{#1}}
\newcommand{\norm}[1]{\|#1\|}
\newcommand{\cY}{\mathcal Y}
\newcommand{\cD}{\mathcal D}
\newcommand{\cC}{\mathcal C}
\newcommand{\cG}{\mathcal G}
\newcommand{\cN}{\mathcal N}
\newcommand{\cA}{\mathcal A}
\newcommand{\cQ}{\mathcal Q}
\newcommand{\cU}{\mathcal U}
\newcommand{\cR}{\mathcal R}
\newcommand{\cH}{\mathcal H}
\newcommand{\cB}{\mathcal B}
\newcommand{\cS}{\mathcal S}
\newcommand{\cI}{\mathcal I}
\newcommand{\R}{\mathds R}
\newcommand{\N}{\mathds N}
\newcommand{\E}{\mathds E}
\renewcommand{\P}{\mathds P}
\newcommand{\bSigma}{\textbf{$\Sigma$}}

@@

This paper introduces a prognostic method called \textit{lights} to deal with the problem of joint modeling of longitudinal data and censored durations, where a large number of both longitudinal and time-independent features are available. 
In the literature, standard joint models are either of type shared random-effect or joint latent class ones ; where the association structure between the longitudinal and the time-to-event submodels takes respectively the form of either shared association features learned from the longitudinal processes and included as potential risk factor in the survival model, or latent classes modeling population heterogeneity.
We pick modeling ideas from both worlds and use appropriate penalties during inference for being able to learn from a high-dimensional context.
The statistical performance of the method is examined on an extensive Monte Carlo simulation study, and finally illustrated on a publicly available dataset.
Our proposed method significantly outperforms the state-of-the-art joint models regarding risk prediction in terms of C-index in a so-called real-time prediction paradigm, with a computing time orders of magnitude faster. In addition, it provides powerful interpretability by automatically pinpointing significant features being relevant from a practical perspective. Thus, we propose a powerfull tool with the ability of automatically determining significant prognostic longitudinal features, which is of increasing importance in many areas: for instance personalized medicine, or churn prediction in a customer profile and activity monitoring setting, to name but a few.\\

@@

In this paper, we propose a method called \textit{lights} (generaLized joInt hiGH-dimensional longiTudinal Survival) which is inspired by both JLCMs and SREMs, since we also include features extracted from the longitudinal processes as potential risk factor in the survival model, which is a group-specific Cox model~\citep{Cox1972JRSS} with high-dimensional shared associations.
To allow flexibility in modeling the dependency between the longitudinal features and the event time, we use appropriate penalties : elastic net~\citep{zou2005regularization}
for time-independent feature selection in the latent class membership, and sparse group lasso~\citep{simon2013sparse} in the survival model : this penalty acts like the lasso at the trajectory level, namely an entire trajectory may drop out of the model, and yields sparsity for a given trajectory.
Inference is achieved using an efficient and novel proximal Quasi-Newton Expectation Maximization algorithm (prox-QNEM).
Posterior estimates of the latent class membership probabilities given the longitudinal process and the fact that the subject is still alive at prediction time -- in a so called ``real-time prediction paradigm'' -- are then used as discriminative marker rule for risk prediction in the cross-validation procedure for selecting the best regularization hyper-parameters.
Hence, the method provides interpretations of the high-dimensional longitudinal features, thus offering a powerful tool for real-time decision support in a customer's satisfaction monitoring context for instance, or else for clinical decision making in patient monitoring.

@@

We suppose a group-specific marker trajectory and a generalized linear mixed model for each longitudinal marker given subgroup $G$, so that for the $l$-th outcome at time $t \geq 0$ one has
\begin{equation}
  \label{eq:link-function}
  h_l\big(\E[Y^l(t)|b^l, G=k]\big) = m_k^l(t)
 \end{equation}
where $h_l$ denotes a known one-to-one link function, and $m_k^l$ the linear predictor such that
\[ m_k^l(t) = u^l(t)^\top\beta_k^l + v^l(t)^\top b^l \]
where $u^l(t) \in \R^{q_l}$ is a row vector of (possibly) time-varying features with corresponding unknown fixed effect parameters $\beta_k^l$, and $v^l(t) \in \R^{r_l}$ is a row vector of (possibly) time-varying features with $r_l \leq q_l$.
A suitable distributional assumption for the random effects component is a zero-mean multivariate normal distribution~\citep{hickey2016joint}, that is
\begin{equation}
   \label{eq:random-effect-law}
   b^l \sim \cN(0, D_{ll})
 \end{equation} 
with $D_{ll} \in \R^{r_l \times r_l}$ the unstructured variance-covariance matrix. 
To account for dependence between the different longitudinal outcome types, we let $\text{Cov}[b^l,b^{l'}] = D_{ll'}$ for $l \ne l'$ and we denote
\[ D = 
\begin{bmatrix}
  D_{11} & \cdots & D_{1L}\\
  \vdots &  \ddots & \vdots \\
  D_{1L}^\top & \cdots & D_{LL}
\end{bmatrix}
\]
the global variance-covariance matrix.
In the sequel of the paper, our aim is to associate the true and unobserved value $m_k^l(t)$ of the $l$-th longitudinal outcome at time $t$ with the event outcome $T^\star$.

@@

To quantify the effect of the longitudinal outcomes on the risk for an event, we use a Cox~\citep{Cox1972JRSS} relative risk model of the form
\begin{equation}
  \label{eq:intensity-model}
  \lambda(t|\cY(t), G = k) = \lambda_0(t) \exp \Big\{\sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l} \Psi_a^l(t) \Big\},
\end{equation}
where $\lambda_0(\cdot)$ is an unspecified baseline hazard function and given that $G = k$, we denote 
\[\cY^l(t) = \{Y^l(u), 0 \leq u < t\} \quad \text{and} \quad \cY(t) = \bigcup_{l=1}^L \cY^l(t)\] 
the history of the true longitudinal process up to time $t$.
For each $l$-th longitudinal outcome, we consider $\cA \in \N_+$ known functionals $\Psi_a^l$ extracted from $\cY^l(t)$ through a given representation mapping, and $\gamma_{k,a}^l \in \R$ the corresponding joint representation parameters.

Let us finally denote
\[\gamma_k= (\gamma_{k,1}^1, \ldots, \gamma_{k,\cA}^1, \ldots, \gamma_{k,1}^L, \ldots, \gamma_{k,\cA}^L )^\top \in \R^{L\cA}, \] so that
\begin{equation}
  \label{eq:def-asso-features}
  \lambda(t|\cY(t), G = k) = \lambda_0(t) \exp \big\{\gamma_k^\top \Psi(t) \big\},
\end{equation}
where $\Psi(t) \in \R^{L\cA}$ denotes the representation features vector.

\subsection{Generalization of SREMs and JLCMs}

Our model is clearly of JLCMs type, and can be viewed as a generalization of SREMs type in a much more flexible way. Table~\ref{table:shared_associations} gives common parameterizations for shared associations that are only considered separately ($\cA=1$) in SREMs.
\begin{table}[htb]
\centering
\begin{tabular}{cccc}
\toprule
Description & $\Psi_a^l(t, \beta_k^l, b^l)$  & Reference \\
\midrule
Linear predictor & $m_k^l(t)$ & \citet{chi2006joint} \\ [.15cm]
Random effects & $b^l$  & \citet{hatfield2011joint} \\
Time-dependent slope & $\dfrac{\dd}{\dd t} m_k^l(t)$ & \citet{rizopoulos2011bayesian} \\ [.3cm]
Cumulative effect & $\int_0^t m_k^l(s) \dd s$ & \citet{andrinopoulou2017combined} \\ [.1cm]
 \bottomrule
\end{tabular}
\caption[]{Examples of classical shared associations in SREMs.}
\label{table:shared_associations}
\end{table}

Our proposition differs from standard SREMs in two respects: $(i)$ representation vector $\Psi(t)$ does not depend on the modeling assumptions in the longitudinal submodel and $(ii)$ it allows to choose multiple high-dimensional representation mappings that characterize time series and concatenate them, since we perform features selection through the regularization strategy described in Section~\ref{sec:penalized-obj}. Point $(i)$ is key since for instance the $\beta_k^l$ dependence in $\Psi_a^l$ would lead to complicated updates, while the $b^l$ dependence would lead to untracktable integrals that are common in SREMs and require approximation methods such as Monte Carlo~\citep{hickey2018joinerml}, being computationally intensive and not scalable in a high-dimensional context. For point $(ii)$ in practice, we use the \texttt{Python} library \texttt{tsfresh}~\citep{christ2018time} and include many extracted features, such as absolute energy of the time series, statistics on autocorrelation, or Fourier and wavelet basis projections, to name but a few.

\subsection{Graphical model}
Figure~\ref{fig:graphical-model} below gives the graphical representation for the lights model to get a beter view of the dependence structure for the model.
\begin{figure}[!htb]
\centering
\resizebox{.9\textwidth}{!}{
\begin{tikzpicture}
\hspace{3cm}
\tikzstyle{main}=[circle, minimum size = 18mm, ultra thick, draw =black!80, node distance = 13mm,text centered]
\tikzstyle{var}=[rectangle, minimum size = 16mm, ultra thick, draw =black!80, node distance = 13mm,text centered]
\tikzstyle{connect}=[-latex, thick]
\tikzstyle{box}=[rectangle, draw=black!100]
  \node[var, fill = white!100] (beta) [] {\Large$\bm{\beta_k^l}$};
  \node[var, fill = white!100] (xi) [below=4cm of beta] {\Large$\bm{\xi_k}$};
  \node[var, fill = white!100] (gamma) [below=3cm of xi] {\Large$\bm{\gamma_k}$};
  \node[main, fill = white!100] (G) [right=4cm of xi] {\Large$\bm{G_i}$};
  \node[main, fill = white!100] (mu) [right=4cm of beta] {\Large$\bm{M_{ik}^l(t)}$};
  \node[main, fill = black!30] (Y) [right=of mu] {\Large$\bm{Y_i^l(t)}$};
  \node[main, fill = white!100] (b) [above=of mu] {\Large$\bm{b_i^l}$};
  \node[var, fill = white!100] (phi) [above=of Y] {\Large$\bm{\phi_l}$};
  \node[var, fill = white!100] (Sigma) [left=of b] {\Large$\bm{D}$};
  \node[main, fill = white!100] (W) [below=3cm of Y] {\Large$\bm{\Psi_i(t)}$};
  \node[main, fill = white!100] (lambda) [right=4cm of gamma] {\Large$\bm{\lambda_i(t)}$};
  \node[main, fill = black!30] (T) [right=of lambda] {\Large$\bm{T_i}$};
  \node[main, fill = black!30] (X) at (3, -8.3) {\Large$\bm{X_i}$};

@@

\end{tikzpicture}}
\caption{Graphical representation of the lights model. Square boxes denote observed data while circles denote unobserved (including random) terms. One can then observe the two links in blue between the submodels that gives rise to the joint model.}
\label{fig:graphical-model}
\end{figure}

@@

Consider an independent and identically distributed (i.i.d.) cohort of $n$ subjects
\[ \cD_n = \big\{ (x_1, y_1^1, \ldots, y_1^L, t_1, \delta_1), \ldots, (x_n, y_n^1, \ldots, y_n^L, t_n, \delta_n) \big\} \]
where for each subject $i=1, \ldots, n$, process $Y_i^l$ is measured $n_i^l$ times at $t_{i1}^l, \ldots, t_{in_i^l}^l$ (which can differ between subjects and outcomes) with $t_{ij}^l \leq t_{ij+1}^l$ for all $j=1, \ldots, n_i^l-1$ and such that
\[ y_i^l=(y_{i1}^l, \ldots, y_{in_i^l}^l)^\top \in \R^{n_i^l} \quad \text{ with } \quad y_{ij}^l=Y_i^l(t_{ij}^l) \]
for all $l=1, \ldots, L$. 
For the $i$-th subject, let us denote 
\[
\left\{
    \begin{array}{ll}
        y_i &= ({y_i^1}^\top \cdots {y_i^L}^\top)^\top \in \R^{n_i},\\
        b_i &= ({b_i^1}^\top \cdots {b_i^L}^\top)^\top \in \R^r,
    \end{array}
\right.
\]
with $n_i = \sum_{l=1}^L n_i^l$ and $r = \sum_{l=1}^L r_l$ the total number of longitudinal measurements (for subject $i$) and the total dimension of the random effects respectively, as well as the following design matrices
\[ U_i = 
\begin{bmatrix}
  U_{i1} & \cdots & 0\\
  \vdots &  \ddots & \vdots \\
  0 & \cdots & U_{iL}
\end{bmatrix} 
\in \R^{n_i \times q}
\qquad \text{and} \qquad
V_i = 
\begin{bmatrix}
  V_{i1} & \cdots & 0\\
  \vdots &  \ddots & \vdots \\
  0 & \cdots & V_{iL}
\end{bmatrix}
\in \R^{n_i \times r}
\]
with $q = \sum_{l=1}^L q_l$ and where for all $l=1, \ldots, L$, one writes
\[
\left\{
    \begin{array}{ll}
        U_{il} &= \big(u_i^l(t_{i1}^l)^\top \cdots u_i^l(t_{in_i^l}^l)^\top\big)^\top \in \R^{n_i^l \times q_l},\\
        V_{il} &= \big(v_i^l(t_{i1}^l)^\top \cdots v_i^l(t_{in_i^l}^l)^\top\big)^\top \in \R^{n_i^l \times r_l}.
    \end{array}
\right.
\]
From now on, all computations are done conditionally on the design matrices $(U_i)_{i=1, \dots, n}$ and $(V_i)_{i=1, \dots, n}$.
Denoting $(t_1^u, \ldots, t_J^u)$ the $J \in \N_+$ unique failure times, we then extract all representation vectors $\big(\Psi_i(t_j^u)\big)_{j=1,\ldots,J}$ for all subjects $i=1, \dots, n$ before inference and from now on, all computations are also done conditionally on them.
Now, let us make two first hypothesis.
\begin{assumption}
\label{assumption1}
Suppose that the random effects are independent of the group membership, and that the latter remain independent conditional on the observed data (namely $T$, $\Delta$ and $Y$).
\end{assumption}
Assumption~\ref{assumption1} states that subject-and-longitudinal outcome specific random effects $b^l$ do not depend on the group membership, which is not a strong modeling assumption.
\begin{assumption}
\label{indep-hyp-2}
Suppose that for a given subject $i$ and conditional on the random effects and the group membership, all $Y_i^l$ are independent for $l=1, \dots, L$.
\end{assumption}
Assumption~\ref{indep-hyp-2} is similar to standard modeling hypothesis, see for instance~\citet{tsiatis2004joint}.
For all $k = 0, \ldots, K-1$, we denote 
\[ \beta_k = ({\beta_k^1}^\top \cdots {\beta_k^L}^\top)^\top \in \R^q \] 
and 
\[M_{ik} = U_i\beta_k + V_ib_i \in \R^{n_i}. \]

@@

Let us denote 
\[ \theta = \big(\xi_0^\top \cdots \xi_{K-1}^\top, \beta_0^\top \cdots \beta_{K-1}^\top, \phi^\top, \text{vech}(D), \lambda_0^\top, \gamma_0^\top \cdots \gamma_{K-1}^\top\big)^\top \in \R^\vartheta \]
the collection of the $\vartheta \in \N_+$ unknown parameters to estimate. Note that we include $\lambda_0$ in $\theta$. Indeed, we will estimate the baseline using nonparametric maximum likelihood, as a function taking mass at each failure time. So in $\theta$, $\lambda_0$ represents the vector of these baseline hazard quantities, with a dimension equal to $J$.
To write the log-likelihood $\ell_n(\theta)$ (rescaled by $n^{-1}$)  for samples in $\cD_n$, corresponding to the joint distribution of the time-to-event and longitudinal outcomes, let us make the following hypothesis.
\begin{assumption}
\label{indep-hyp-3}
Assume that
\begin{equation}
  \label{eq:ind-hyp}
  f_\theta(t_i, \delta_i, y_i | b_i, G_i = k) = f_\theta(t_i, \delta_i| G_i = k) f_\theta(y_i | b_i, G_i = k)
\end{equation}
for all $i=1, \ldots, n$.
\end{assumption}
Assumption~\ref{indep-hyp-3} is a generalization of classical hypothesis used in SREMs and JLCMs~\citep{hickey2016joint}. Remember that the densities are also conditionned by all the representation features $\big(\Psi_i(t_j^u)\big)_{j=1,\ldots,J}$. Hence Assumption~\ref{indep-hyp-3} says that the random effects vector $b_i$, the group membership and the representation features account for the association between the longitudinal and event outcomes, which makes sense given Figure~\ref{fig:graphical-model}.
Then, one has
\begin{equation}
	\label{eq:log-lik}
	\ell_n(\theta) = \ell_n(\theta ; \cD_n) = n^{-1} \sum_{i=1}^n \log \sum_{k=0}^{K-1} \pi_{\xi_k}(x_i) f_\theta(t_i, \delta_i| G_i = k) f_\theta(y_i | G_i = k),
\end{equation}
where 
\[f_\theta(t_i, \delta_i| G_i = k) = \big[\lambda(t_i|\cY_i(t_i), G_i = k)\big]^{\delta_i} \exp \Big\{-\int_0^{t_i} \lambda(s|\cY_i(s), G_i = k) \dd s \Big\}. \]
As expected, one has a JLCMs like log-likelihood.

@@

In many practical applications, subjects show highly nonlinear longitudinal trajectories. We consider a flexible representations for $u^l(t)$ using a high-dimensional vector of time monomials, namely 
\begin{equation}
  \label{eq:alpha-def}
  u^l(t) = (1, t, t^2, \ldots, t^\alpha)^\top
\end{equation}
with $\alpha \in \N_+$.
The idea here is to let the practitioner choose a suitable polynomial order $\alpha$ for the representation -- that could also be tuned automatically using a model selection procedure and AIC~\citep{akaike1974new} for instance.
We then let \[v^l(t) = (1, t)^\top\] so that each trajectory of each subject gets an affine random effect. Hence with this choice in practice, one has $q_l=\alpha + 1$ and $r_l=2$ for all $l=1, \ldots, L$.

@@

\subsection{A Proximal Quasi-Newton EM}
\label{sec:prox-QNEM}
